Pigeons.jl: Distributed sampling from intractable distributions

### 2024-03-23 paper/PT.tex
use paper folder
First updates based on reviewer comments

\begin{figure}[t]
    \centering
    \includegraphics[width=0.3\textwidth]{img/bimodal.pdf}
    \caption{A simple bimodal distribution from which traditional 
    MCMC methods may struggle to obtain samples. Blue lines display output 
    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
    The sampler in this figure is visibly trapped in one of the two modes.}
    \label{fig:bimodal}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{subfigure}
    \begin{subfigure}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{subfigure}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

@@

\begin{figure*}[t] 
  \centering 
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_1_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_1.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_2_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_2.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_3_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_3.pdf}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{img/Communication_4_v3.pdf}
    \includegraphics[width=0.59\textwidth]{img/index_process_4.pdf}
  \end{subfigure}
  \caption{
    A summary of communication between four machines with $N=4$ PT chains.
    Circles represent machines with the machine number in the center of the circle. 
    Different colours are also used for different machines.
    Numbers in parantheses represent the \texttt{chain} 
    that is currently being explored by that machine. 
    Arrows indicate communication occuring between machines to exchange \texttt{chain} 
    indices.
    To keep track of which chains are stored on which machine, we introduce the 
    \texttt{dist\_array} of type \texttt{PermutedDistributedArray}, described in 
    \cref{sec:permuted_dist_array}, which is of length $N$. 
    The $j$-th element of \texttt{dist\_array} at a given time step indicates which 
    machine number is storing \texttt{chain} index $j$.
    The bolded curve in the figures to the right indicate the trajectory of the first 
    \texttt{Replica} over the course of each of the communication steps.      
    \textbf{Top left:} the PT replicas are initialized and the first communication step 
    is proposed. At this time step, $\texttt{dist\_array} = [1,2,3,4]$.
    \textbf{Top right:} after the first successful communication step, we now have 
    $\texttt{dist\_array} = [2,1,4,3]$.
    \textbf{Bottom left:} The second communication step is completed and 
    $\texttt{dist\_array} = [2,4,1,3]$.
    \textbf{Bottom right:} The third communication step is completed and 
    $\texttt{dist\_array} = [4,2,3,1]$.}
  \label{fig:index_process}
\end{figure*}

@@

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=0.3\textwidth]{../img/bimodal.pdf}
    \caption{A simple bimodal distribution from which traditional 
    MCMC methods may struggle to obtain samples. Blue lines display output 
    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
    The sampler in this figure is visibly trapped in one of the two modes.}
    \label{fig:bimodal}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{minipage}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

@@

\begin{figure*}[t] 
  \centering 
  \begin{minipage}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{../img/Communication_1_v3.pdf}
    \includegraphics[width=0.59\textwidth]{../img/index_process_1.pdf}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{../img/Communication_2_v3.pdf}
    \includegraphics[width=0.59\textwidth]{../img/index_process_2.pdf}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{../img/Communication_3_v3.pdf}
    \includegraphics[width=0.59\textwidth]{../img/index_process_3.pdf}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
    \centering
    \includegraphics[width=0.39\textwidth]{../img/Communication_4_v3.pdf}
    \includegraphics[width=0.59\textwidth]{../img/index_process_4.pdf}
  \end{minipage}
  \caption{
    A summary of communication between four machines with $N=4$ PT chains.
    Circles represent machines with the machine number in the center of the circle. 
    Different colours are also used for different machines.
    Numbers in parantheses represent the \texttt{chain} 
    that is currently being explored by that machine. 
    Arrows indicate communication occuring between machines to exchange \texttt{chain} 
    indices.
    To keep track of which chains are stored on which machine, we introduce the 
    \texttt{dist\_array} of type \texttt{PermutedDistributedArray}, described in 
    \cref{sec:permuted_dist_array}, which is of length $N$. 
    The $j$-th element of \texttt{dist\_array} at a given time step indicates which 
    machine number is storing \texttt{chain} index $j$.
    The bolded curve in the figures to the right indicate the trajectory of the first 
    \texttt{Replica} over the course of each of the communication steps.      
    \textbf{Top left:} the PT replicas are initialized and the first communication step 
    is proposed. At this time step, $\texttt{dist\_array} = [1,2,3,4]$.
    \textbf{Top right:} after the first successful communication step, we now have 
    $\texttt{dist\_array} = [2,1,4,3]$.
    \textbf{Bottom left:} The second communication step is completed and 
    $\texttt{dist\_array} = [2,4,1,3]$.
    \textbf{Bottom right:} The third communication step is completed and 
    $\texttt{dist\_array} = [4,2,3,1]$.}
  \label{fig:index_process}
\end{figure*}

@@

\subsection{Strong and weak scaling of PT}

We study the strong and weak scaling properties of Pigeons' implementation of PT
as given by Amdahl's law and Gustafson's law, respectively.
For the strong scaling results, we consider a fixed problem size and compare the 
compute time as we vary the number of processes.  
For weak scaling, we consider a varying problem size proportional 
to the number of processes used. 
In both cases we use the number of chains as the definition of the problem size; 
this is in line with previous heuristics that establish that the generaly problem 
difficulty is proportional to the optimal number of chains in PT \cite{syed2021nrpt}.

\medskip
We emphasize again that Pigeons supports both shared memory 
parallelism (SMP) and distributed memory parallelism (DMP). However, in certain cases, it 
might be useful to run parallel tempering with very many chains or 
to run several instances of parallel tempering at once. In such cases, 
the DMP setup is the most natural. Our implementation in Pigeons ensures that the 
communication cost in this setup is minimal. In most cases, the 
communication cost is negligible compared to other computations performed on each process.

\medskip
The results for strong and weak scaling are presented in \cref{fig:PT_scaling}. 
For strong scaling, we see that ...
For weak scaling, We see that ...
\NS{Comment on plots}.

\begin{figure*}[t]
    \centering
    \begin{minipage}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
    \end{minipage}
    \caption{
        \NS{Add appropriate plots.}
        Strong and weak scaling results for PT.  
        \textbf{Left:} Strong scaling. 
        \textbf{Right:} Weak scaling. 
    }
    \label{fig:PT_scaling}
\end{figure*}

### 2024-03-23 paper/introduction.tex
First updates based on reviewer comments

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
enables users to sample efficiently from high-dimensional and complex distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple API allows users to perform such computations 
single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
machines. Further, it comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

>>>>

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
enables users to sample efficiently from high-dimensional and complex distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple API allows users to perform such computations 
single-threaded, multi-threaded, and/or distributed over thousands of MPI-communicating 
machines (as tested on local compute clusters). Further, it comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}
